---
title: "Microbiome_rf"
author: "Chiranjit Mukherjee"
date: "10/18/2020"
output: html_document
  keep_md=TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("randomForest")
library("plyr") # for the "arrange" function
library("rfUtilities") # to test model significance
library("caret") # to get leave-one-out cross-validation accuracies and also contains the nearZeroVar function
library(e1071)
library(ggplot2)
library(vegan)
```

```{r}
otu_table <- read.table("otu_table_RF_tutorial.txt", sep="\t", header=T, row.names=1, stringsAsFactors=FALSE, comment.char="")  
metadata <- read.table("metadata_RF_tutorial.txt", sep="\t", header=T, row.names=1, stringsAsFactors=TRUE, comment.char="")
```

```{r}
dim(otu_table)
#[1] 1000   40

dim(metadata)
#[1] 40  2

str(metadata)
```

```{r}
otu_table_pct <- data.frame(t(otu_table), check.names = FALSE)

# Compute NMDS
set.seed(12345); capture.output(otu_table_pct.mds <- metaMDS(otu_table_pct, trymax = 200, autotransform = F, wascores = F))

# Permanova
set.seed(12345); otu_table_pct.perm <- adonis(formula = otu_table_pct ~ metadata$state) # R2 0.19607, p < 0.001

# Making dataframe for plotting
otu_table_pct.mds.df <- data.frame(scores(otu_table_pct.mds, display = 'sites'))
otu_table_pct.mds.df$state <- metadata$state[match(row.names(otu_table_pct.mds.df), row.names(metadata))]

# Colors
two_cols <- c("forestgreen", "firebrick4")
```

```{r}
# Plot NMDS
ggplot(otu_table_pct.mds.df, aes(x=NMDS1, y=NMDS2)) + stat_ellipse(alpha=0.8, aes(color=state), show.legend = F, lwd=0.2) + 
  geom_point(alpha=0.9, aes(fill = state), size=3, color="black", pch=21, stroke=0.2) + scale_fill_manual(values=two_cols) + scale_color_manual(values=two_cols) + 
  labs(title = "Species Level: Beta Diversity Comparison", subtitle = "Relative Abundance", fill="Subject state") +
  annotate("text", x = max(otu_table_pct.mds.df$NMDS1)-0.2, y = max(otu_table_pct.mds.df$NMDS2), 
           label = paste("p <", otu_table_pct.perm$aov.tab$`Pr(>F)`[1], "(PERMANOVA)", sep=" ")) + theme_classic() +
  theme(plot.title = element_text(size=15, face="bold", hjust=0.5), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", axis.line = element_line(size = 0.3),
        legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
ggsave(file = "figs/otu_table_pct_mds.pdf", width = 10, height = 6, units = "in")

```


#### Filter out rare OTUs 
Keeping only those OTUs which have less than 30 (75% of total number of samples) times 0 values in the dataset
```{r}
otu_table_rare_removed <- otu_table[rowSums(otu_table == 0) <= 0.75*ncol(otu_table), ]
dim(otu_table_rare_removed) #  [1] 595  40
```


#### Re-normalize_table so that each sample's column sums to 1
```{r}
otu_table_rare_removed_norm <- sweep(otu_table_rare_removed, 2, colSums(otu_table_rare_removed) , '/')*1
colSums(otu_table_rare_removed_norm)
```


```{r}
otu_table_rare_removed_norm_pct <- data.frame(t(otu_table_rare_removed_norm), check.names = FALSE)

# Compute NMDS
set.seed(12345); capture.output(otu_table_rare_removed_norm_pct.mds <- metaMDS(otu_table_rare_removed_norm_pct, trymax = 200, autotransform = F, wascores = F))

# Permanova
set.seed(12345); otu_table_rare_removed_norm_pct.perm <- adonis(formula = otu_table_rare_removed_norm_pct ~ metadata$state) # R2 0.19607, p < 0.001

# Making dataframe for plotting
otu_table_rare_removed_norm_pct.mds.df <- data.frame(scores(otu_table_rare_removed_norm_pct.mds, display = 'sites'))
otu_table_rare_removed_norm_pct.mds.df$state <- metadata$state[match(row.names(otu_table_rare_removed_norm_pct.mds.df), row.names(metadata))]

# Colors
two_cols <- c("forestgreen", "firebrick4")
```

```{r}
# Plot NMDS
ggplot(otu_table_rare_removed_norm_pct.mds.df, aes(x=NMDS1, y=NMDS2)) + stat_ellipse(alpha=0.8, aes(color=state), show.legend = F, lwd=0.2) + 
  geom_point(alpha=0.9, aes(fill = state), size=3, color="black", pch=21, stroke=0.2) + scale_fill_manual(values=two_cols) + scale_color_manual(values=two_cols) + 
  labs(title = "Species Level: Beta Diversity Comparison", subtitle = "Relative Abundance", fill="Subject state") +
  annotate("text", x = max(otu_table_rare_removed_norm_pct.mds.df$NMDS1)-0.2, y = max(otu_table_rare_removed_norm_pct.mds.df$NMDS2), 
           label = paste("p <", otu_table_rare_removed_norm_pct.perm$aov.tab$`Pr(>F)`[1], "(PERMANOVA)", sep=" ")) + theme_classic() +
  theme(plot.title = element_text(size=15, face="bold", hjust=0.5), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", axis.line = element_line(size = 0.3),
        legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
#ggsave(file = "otu_table_rare_removed_norm_pct_mds.pdf", width = 10, height = 6, units = "in")

```




#### Data Transformation
One approach is to standardize the data by subtracting each sample's mean (center) and then dividing by the sample's standard deviation (scale). In other words, each value is converted into a Z-score.
```{r}
otu_table_scaled <- scale(otu_table_rare_removed_norm, center = TRUE, scale = TRUE)  
```
Negative counts values!





#### Running Model
input tables for classification of state:
```{r}
otu_table_scaled_state <- data.frame(t(otu_table_scaled))  
otu_table_scaled_state$state <- metadata[rownames(otu_table_scaled_state), "state"]  
dim(otu_table_scaled_state)
```


input tables for regression of inflammation score (IS):
```{r}
otu_table_scaled_IS <- data.frame(t(otu_table_scaled))  
otu_table_scaled_IS$IS <- metadata[rownames(otu_table_scaled_IS), "IS"]
```



```{r}
set.seed(123); RF_state_classify <- randomForest(x=otu_table_scaled_state[, 1:(ncol(otu_table_scaled_state)-1)],
                                                 y=otu_table_scaled_state[, ncol(otu_table_scaled_state)],
                                                 ntree=501, importance=TRUE, proximities=TRUE )

set.seed(123); RF_IS_regress <- randomForest(x=otu_table_scaled_IS[, 1:(ncol(otu_table_scaled_IS)-1)], 
                                             y=otu_table_scaled_IS[, ncol(otu_table_scaled_IS)],
                                             ntree=501, importance=TRUE, proximities=TRUE )  
```


### Permutation Testing Model Performance

```{r, eval=FALSE}
#Evaluate is set to false to reduce Knit time
RF_state_classify_sig <- rf.significance( x=RF_state_classify,
                                          xdata=otu_table_scaled_state[,1:(ncol(otu_table_scaled_state)-1)],
                                          nperm=1000 , ntree=501 )  


RF_IS_regress_sig <- rf.significance( x=RF_IS_regress,
                                      xdata=otu_table_scaled_IS[,1:(ncol(otu_table_scaled_IS)-1)],
                                      nperm=1000 , ntree=501 )  
```


### Accuracy Estimated by Cross-validation

```{r}
fit_control <- trainControl(method = "LOOCV")    
RF_state_classify_loocv <- train(otu_table_scaled_state[,1:(ncol(otu_table_scaled_state)-1)],
                                  y=otu_table_scaled_state[, ncol(otu_table_scaled_state)], 
                                  method="rf", ntree=501 , tuneGrid=data.frame( mtry=25 ), trControl=fit_control)

RF_IS_regress_loocv <- train(otu_table_scaled_IS[,1:(ncol(otu_table_scaled_IS)-1)],
                             y=otu_table_scaled_IS[, ncol(otu_table_scaled_IS)], 
                             method="rf", ntree=501 , tuneGrid=data.frame( mtry=215 ), trControl=fit_control)

```



## Identifying Important Features

```{r}
RF_state_classify_imp <- as.data.frame(RF_state_classify$importance)
RF_state_classify_imp$features <- rownames(RF_state_classify_imp)
RF_state_classify_imp_sorted <- arrange(RF_state_classify_imp, desc(MeanDecreaseAccuracy))

barplot(RF_state_classify_imp_sorted[1:10,"MeanDecreaseAccuracy"], names.arg=RF_state_classify_imp_sorted[1:10,"features"] , 
        ylab="Mean Decrease in Accuracy (Variable Importance)", las=2, ylim=c(0,0.02), main="Classification RF") 
```


```{r}


ggplot(RF_state_classify_imp_sorted[c(1:25), ], aes(features, MeanDecreaseAccuracy)) + geom_bar(stat="identity")
```



#### Before and After NMDS 


